---
title: "Assignment 4: Data Visualization, Git, and Quarto"
author: "Akbar Naqvi and Roujia Lin"
format: html
editor: visual
---

## Data Used

For this assignment, we are using the **Community Resilience Estimates from the U.S Census Bureau:** (<https://www.census.gov/programs-surveys/community-resilience-estimates.html>)

The CRE tracks how at-risk every single neighborhood in the United States is to the impacts of COVID-19 and other local disasters, by measuring the capacity of individuals and households at absorbing, enduring and recovering from the external stresses of the impacts of a disaster.

## Figure 1

This scatterplot maps the relationship between the percentage of households that are at high risk for a disaster, defined by the census as having 3 or more risk factors (defined [here](https://www.census.gov/programs-surveys/community-resilience-estimates/technical-documentation/methodology.html)),that make them vulnerable to a disaster and the percentage of that population's state that lives in a rural area, and adds an additional layer by demonstrating the population of the state using the size of the bubbles.

```{r}

library(tidyverse)
cre_equity_state<-read_csv("https://www2.census.gov/programs-surveys/demo/datasets/community-resilience/2019/CRE_Equity_State_19.csv")


ggplot(data = cre_equity_state,
       mapping = aes(x = Rural_Pop_PE,
                     y = PRED3_PE,
                     size = Total_Population_E)) + 
  geom_point(color = "red", alpha = 0.7)+
  scale_size(range = c(.1,15))+
  labs(title = paste("Disaster Risk vs. Rurality"),
       subtitle = paste("Both highly urban and highly rural populations can be vulnerable to disasters",
                        "are less resilient to disasters"),
       caption = paste("Source: U.S Census Bureau Community Resilience Estimates"),
       size = "Total Population",
       x = "Percent of Population Living in Rural Areas ",
       y = "% people with 3+ risk factors
       \nfor vulnerability to a disaster", color = "Total Population") +
   geom_text(aes(label = NAME), 
            show.legend = F,
            size = 1.75)+
  theme(legend.position = "none")


```

As would be expected, larger states tend to have a smaller share of the population living in rural areas. What is interesting about this figure is that we see absolutely no relationship whatsoever between the percent of people who are "high risk" for a disaster and the percent of people who live in a rural area. Given that rural areas have less access to broadband and experience high levels of poverty, both inputs into the CRE, it is surprising to see that a state's share of people living in rural areas does not have an impact on the share of people experiencing a disaster. This might demonstrate that people in urban areas face other challenges unique to them that also make them vulnerable to disasters.

## Figure 2

```{r,Minority Community Resilience}


cre_equity_county<-read_csv("https://www2.census.gov/programs-surveys/demo/datasets/community-resilience/2019/CRE_Equity_County_19.csv")

#Plot High Risk Majority-Black Counties

cre_equity_county %>%
  mutate("more_white" = if_else(NH_White_alone_PF ==1|NH_White_alone_PF == 0 , "Less Non-White", "More Non_White")) %>%
  filter(PRED3_PF == 1) %>%
  ggplot() +
  geom_bar(mapping = aes(x = more_white), 
           alpha = 0.2,
           fill = "blue") +
  theme_classic()+
  labs(title = paste("Minority Community Resilience"),
       subtitle = paste("Communities with a higher share of minorities are less resilient than those with a lower share"),
       caption = paste("Source: U.S Census Bureau"),
       x = "Non-White Share Compared to the National Average",
       y = "Number of Counties")


```

This graph describes the difference between the resilience of communities with a high concentration of minorities. We see that a larger number of counties that are more non-white than average have a higher percentage of individuals that are described as "high risk" by the community resilience index. This confirms the hypothesis that areas with more people of color have more people that are considerably vulnerable to the effects of a disaster, something reflected in the disproportionate impact of the Covid-19 pandemic on minorities.

## Figure 3

```{r}



#Plot Gini Index vs single mother households

ggplot(data = cre_equity_state,

       mapping = aes(x =Female_no_partner_w_child_PE,y = GINI_IND_Inequality_E))+

  geom_smooth(method = "lm",
              color = "red")+

  geom_point()+

scale_x_continuous(breaks = seq(0,8, by = 1), 
                     limits = c(3,8))+

  labs(title = paste("Inequality in Single-Mother Households"),

       subtitle = paste("States with more single mothers experience more inequality"),

       caption = paste("Source: U.S Census Bureau"),

       x = "% of Households With a Single Mother",

       y = "Gini Index of Inequality")

```

From this graph we key in on a major risk factor in the CRE: the presence of a single caregiver, in this case a single mother. If the risk factor is valid, we would expect that regions with more single mother households experience higher levels of inequality, which are usually associated with high levels of instability, rendering individuals in that region vulnerable to disasters. In general, we see that when the percentage of the single mother households is associated with higher income inequality.

## Figure 4

```{r, Gini Index vs difference between the population without health insurance and national estimate}
#\| warning: false #Read the package library(tidyverse) library(ggplot2)

#Read the dataset cre_equity_state \<- read_csv("https://www2.census.gov/programs-surveys/demo/datasets/community-resilience/2019/CRE_Equity_State_19.csv")

#Plot the density of the difference between the population without health insurance and national estimate 
ggplot(data = cre_equity_state, mapping = aes(x =No_Health_Ins_PF))+
  geom_density(alpha = 1,fill="White",color = "grey")+ 
  scale_size(range = c(.1,10))+ 
  labs(title = paste("population without health insurance vs national estimate"),
       subtitle = paste("the differnce between the population without health insurance and national estimate"), 
       caption = paste("Source: U.S Census Bureau"), 
       size = "Total Population", 
       x = "the difference between the population who do not have health insurance and natioanl estimate", 
       y = "density", color = "Total Population") 

```

variable x: Flag denoting if the estimated percentage of civilian noninstitutionalized population with no health insurance coverage is significantly different than the national estimate.

In nearly 60% of all states, among the people who do not have health insurance, the estimate is lower than the national estimate and it is statistically different.In nearly 38% of all states,among the people who do not have health insurance, the estimate is higher than the national estimate and it is statistically different.

[extract](%22https://educationdata.urban.org/data-explorer/explorer)
